Multilevel constraint decomposition methods \\ for nonlinear variational inequalities

### 2022-10-16 paper/mcd2.tex
a few more edits

We will solve the same VI problem as the abstract problem \eqref{eq:vi}.  Given continuous $f:\cK \to \cV'$ and $\ell \in \cV'$, we assume that the following VI is well-posed for $u^*\in \cK$:
\begin{equation}
\ip{f(u^*)}{v-u^*} \ge \ip{\ell}{v-u^*} \qquad \text{for all } v\in \cK. \label{eq:boxdirichletvi}
\end{equation}
Note that $v-u^* \in W_0^{1,p}(\Omega)$, that is, the test functions in \eqref{eq:boxdirichletvi} have zero values on the Dirichlet boundary $\partial_D\Omega$.  In general, problem \eqref{eq:boxdirichletvi} is a mixed boundary value problem with homogeneous Neumann, i.e.~natural, boundary conditions over $\partial_N \Omega = \Omega \setminus \partial_D \Omega$.

@@

While \eqref{eq:fe:vi} approximates \eqref{eq:boxdirichletvi}, it may not technically be conforming for that VI.  That is, even if quadrature and other implementation details for $f^J$ are ignored, and despite the vector space inclusion $\mathcal{V}^J \subset \mathcal{V}$, we do not require $\mathcal{K}^J \subset \mathcal{K}$.  The continuum constraint set $\mathcal{K}$ remains abstract in applications where only the nodal obstacles $\underline{\gamma}^J, \overline{\gamma}^J$ are defined.  In fact, even if smooth continuum obstacles exist, admissibility with respect to their FE interpolants might not imply continuum admissiblity.\footnote{Such caveats apply in geophysical situations where the obstacle is given by meshed elevation data, the continuum limit of which is undefined, for example \cite{Bueler2016}.}

@@

FIXME READ FROM HERE Finally, the monotone restrictions $\maxR,\minR$ are nonlinear maps which maximize and minimize, respectively, nodal values over the supports of coarser-mesh hat functions.  To make the definitions precise, first denote the \emph{nodes} of mesh $\mathcal{T}^j$ by $x_q^j$ for $q=0,\dots,m_j-1$, and then the corresponding hat functions \cite{Elmanetal2014} by $\psi_q^j$.  (Boundary nodes are included here.)  The monotone restrictions of $z\in\tilde{\mathcal{V}}^j$ are
\begin{align}
\maxR z &= \sum_{p=1}^{\tilde m_{j-1}} \max \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}\, \psi_p^{j-1}, \label{eq:fe:monotonerestrictions} \\
\minR z &= \sum_{p=1}^{\tilde m_{j-1}} \min \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}\, \psi_p^{j-1}, \notag
\end{align}
with the result in $\tilde{\mathcal{V}}^{j-1}$.  The usual ordering rules for the extended real line $\tilde\RR$ are applied.  Note that if $z\in\tilde{\mathcal{V}}^j$ then $\minR z \le z \le \maxR z$.

To define the constraint subsets, the sets generically denoted $\cK_i$ in Section \ref{sec:cd}, in the current multilevel FE context, suppose $w^J \in \cK^J$ is an admissible fine-level iterate which approximates the solution $u^J$ of VI \eqref{eq:fe:vi}.  From this iterate we define the finest-level \emph{lower and upper defect constraints} \cite{GraeserKornhuber2009}
\begin{equation}
\underline{\chi}^J = \underline{\gamma}^J - w^J, \qquad \overline{\chi}^J = \overline{\gamma}^J - w^J, \label{eq:fe:defectconstraints}
\end{equation}
functions in $\tilde{\mathcal{V}}^J$ satisfying $\underline{\chi}^J \le 0 \le \overline{\chi}^J$.  The motivation here is that a corrected iterate is admissible, $\underline{\gamma}^J \le w^J + z^J \le \overline{\gamma}^J$, if and only if the perturbation $z^J$ is between the defect constraints, $\underline{\chi}^J \le z^J \le \overline{\chi}^J$.

For $j=J,J-1,\dots,1$ we apply monotone restrictions to define the \emph{level defect constraints},
\begin{equation}
\underline{\chi}^{j-1} = \maxR \underline{\chi}^j, \qquad \overline{\chi}^{j-1} = \minR \overline{\chi}^j. \label{eq:fe:chilevels}
\end{equation}
Observe that
\begin{equation}
\underline{\chi}^{J} \le \dots \le \underline{\chi}^0 \le 0 \le \overline{\chi}^0 \le \dots \le \overline{\chi}^J. \label{eq:fe:chiordering}
\end{equation}
A $J=4$ example, with no upper obstacles ($\overline{\chi}^j=+\infty$), is shown in Figure \ref{fig:chiphilevels}.

@@

We also compute the differences between the level defect constraints:
\begin{equation}
\underline{\phi}^j = \underline{\chi}^j - \underline{\chi}^{j-1}, \qquad \overline{\phi}^j = \overline{\chi}^j - \overline{\chi}^{j-1},  \label{eq:fe:philevels}
\end{equation}
using $\underline{\chi}^{-1}=\overline{\chi}^{-1}=0$.  Again these constraints bracket zero,
\begin{equation}
\underline{\phi}^j \le 0 \le \overline{\phi}^j,  \label{eq:fe:phibrackets}
\end{equation}
but these functions are not ordered as in \eqref{eq:fe:chiordering}, but telescoping sums hold for $j=0,1,\dots,J$:
\begin{equation}
\sum_{i=0}^j \underline{\phi}^i = \underline{\chi}^j, \qquad \sum_{i=0}^j \overline{\phi}^i = \overline{\chi}^j.  \label{eq:fe:telescoping}
\end{equation}

A multilevel V-cycle is defined in the next Section.  For the downward direction we will use the differences $\underline{\phi}^j,\overline{\phi}^j$ as constraints, and for the upward part the defect constraints $\underline{\chi}^j,\overline{\chi}^j$.  That is,
\begin{align}
\mathcal{D}^j = \{v \in \mathcal{V}^j \,:\, \underline{\phi}^j \le v \le \overline{\phi}^j\}, \label{eq:fe:constraintsets} \\
\mathcal{U}^j = \{v \in \mathcal{V}^j \,:\, \underline{\chi}^j \le v \le \overline{\chi}^j\}, \notag
\end{align}
are the \emph{downward} and \emph{upward constraint subsets} of the FE subspaces $\mathcal{V}^j$, respectively.  Observe that the upward sets are larger, $\mathcal{D}^j \subset \mathcal{U}^j$, and in fact by \eqref{eq:fe:telescoping} the set decompositions
\begin{equation}
\mathcal{U}^j = \sum_{i=0}^j \mathcal{D}^i \label{eq:fe:constraintdecomp}
\end{equation}
apply in the same sense as \eqref{eq:constraintdecomp}.  Also note that the finest-level upward constraint set is equivalent to the original constraint set $\mathcal{K}^J$; for all $z^J \in \mathcal{V}^J$,
\begin{equation}
z^J \in \mathcal{U}^J \quad \iff \quad w^J+z^J \in \mathcal{K}^J. \label{eq:fe:finestlevelequivalent}
\end{equation}

>>>>

We will solve the same VI problem as the abstract problem \eqref{eq:vi}.  Given continuous $f:\cK \to \cV'$ and $\ell \in \cV'$, we assume that the following VI is well-posed for $u^*\in \cK$:
\begin{equation}
\ip{f(u^*)}{v-u^*} \ge \ip{\ell}{v-u^*} \qquad \text{for all } v\in \cK. \label{eq:boxdirichletvi}
\end{equation}
Note that $v-u^* \in W_0^{1,p}(\Omega)$, that is, the test functions in \eqref{eq:boxdirichletvi} have zero values on the Dirichlet boundary $\partial_D\Omega$.

Problem \eqref{eq:boxdirichletvi} is a mixed boundary value problem, in weak form, with Dirichlet conditions $u=g_D$ over $\partial_D\Omega$ and Neumann boundary conditions over $\partial_N \Omega = \Omega \setminus \partial_D \Omega$.  The Neumann boundary conditions are determined by the formula for $f$.  (That is, nonhomogeneous Neumann conditions imply that the nonlinear operator $f:\cK\to\cV'$ involves an integration over $\partial_N\Omega$.)

@@

While \eqref{eq:fe:vi} approximates \eqref{eq:boxdirichletvi}, it may not technically be conforming for that VI.  That is, even if quadrature and other implementation details for $f^J$ are ignored, and despite the vector space inclusion $\mathcal{V}^J \subset \mathcal{V}$, we do not require $\mathcal{K}^J \subset \mathcal{K}$.  The continuum constraint set $\mathcal{K}$ remains abstract in applications where only the nodal obstacles $\underline{\gamma}^J, \overline{\gamma}^J$ are defined.  In fact, even if smooth continuum obstacles exist, admissibility with respect to their FE interpolants might not imply continuum admissiblity.\footnote{Such caveats apply in geophysical situations where the obstacle is given by meshed elevation data, the continuum limit of which is undefined \cite{Bueler2016}.}

@@

Finally, the monotone restrictions $\maxR,\minR$ are nonlinear maps which maximize and minimize, respectively, nodal values over the supports of coarser-mesh hat functions.  To make the definitions precise, first denote the \emph{nodes} of mesh $\mathcal{T}^j$ by $x_q^j$ for $q=0,1,\dots,m_j-1$, and then the corresponding hat functions \cite{Elmanetal2014} by $\psi_q^j(x)$.  The monotone restrictions $\maxR z,\minR z$ of $z\in\tilde{\mathcal{V}}^j$ are the coarser-level functions in $\tilde{\mathcal{V}}^{j-1}$ defined by
\begin{align}
(\maxR z)(x_p^{j-1}) &= \max \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}, \label{eq:fe:monotonerestrictions} \\
(\minR z)(x_p^{j-1}) &= \min \{z(x_q^j) \,:\, \psi_p^{j-1}(x_q^j) > 0\}, \notag
\end{align}
for $p=0,1,\dots,m_{j-1}-1$.  In computing $\max$ and $\min$, the usual ordering rules for the extended real line $\tilde\RR$ are applied.  Note that $\minR z \le z \le \maxR z$ because, by mesh refinement, node $x_p^{j-1}$ is in the set of $x_q^j$ such that $\psi_p^{j-1}(x_q^j)>0$.

To define the constraint subsets in the current multilevel FE context, namely the sets denoted $\cK_i$ in Section \ref{sec:cd}, suppose $w^J \in \cK^J$ is an admissible finest-level iterate which approximates the solution $u^J$ of VI \eqref{eq:fe:vi}.  From this iterate we define the finest-level \emph{lower and upper defect constraints} \cite{GraeserKornhuber2009}
\begin{equation}
\underline{\chi}^J = \underline{\gamma}^J - w^J, \qquad \overline{\chi}^J = \overline{\gamma}^J - w^J, \label{eq:fe:defectconstraints}
\end{equation}
functions in $\tilde{\mathcal{V}}^J$ satisfying $\underline{\chi}^J \le 0 \le \overline{\chi}^J$.  FIXME ADD DIRICHLET The motivation here is that a corrected iterate is admissible, $\underline{\gamma}^J \le w^J + z^J \le \overline{\gamma}^J$, if and only if the perturbation $z^J$ is between the defect constraints, $\underline{\chi}^J \le z^J \le \overline{\chi}^J$.

For $j=J,J-1,\dots,1$ we apply monotone restrictions to define the \emph{level defect constraints},
\begin{equation}
\underline{\chi}^{j-1} = \maxR \underline{\chi}^j, \qquad \overline{\chi}^{j-1} = \minR \overline{\chi}^j. \label{eq:fe:chilevels}
\end{equation}
Observe that
\begin{equation}
\underline{\chi}^{J} \le \dots \le \underline{\chi}^0 \le 0 \le \overline{\chi}^0 \le \dots \le \overline{\chi}^J. \label{eq:fe:chiordering}
\end{equation}
A $J=4$ example, with no upper obstacles, namely $\overline{\chi}^j=+\infty$ for all $j$, is shown in Figure \ref{fig:chiphilevels}.

@@

We also compute the differences between the level defect constraints:
\begin{equation}
\underline{\phi}^j = \underline{\chi}^j - \underline{\chi}^{j-1}, \qquad \overline{\phi}^j = \overline{\chi}^j - \overline{\chi}^{j-1},  \label{eq:fe:philevels}
\end{equation}
using $\underline{\chi}^{-1}=\overline{\chi}^{-1}=0$ to define $\underline{\phi}^0=\underline{\chi}^0$ and $\overline{\phi}^0=\overline{\chi}^0$.  These constraints bracket zero,
\begin{equation}
\underline{\phi}^j \le 0 \le \overline{\phi}^j.  \label{eq:fe:phibrackets}
\end{equation}
Though these functions are not ordered as in \eqref{eq:fe:chiordering}, telescoping sums hold for $j=0,1,\dots,J$:
\begin{equation}
\sum_{i=0}^j \underline{\phi}^i = \underline{\chi}^j, \qquad \sum_{i=0}^j \overline{\phi}^i = \overline{\chi}^j.  \label{eq:fe:telescoping}
\end{equation}

FIXME A multilevel V-cycle is defined in the next Section.  For the downward direction we will use the differences $\underline{\phi}^j,\overline{\phi}^j$ as constraints, and for the upward part the defect constraints $\underline{\chi}^j,\overline{\chi}^j$.  That is,
\begin{align}
\mathcal{D}^j = \{v \in \mathcal{V}^j \,:\, \underline{\phi}^j \le v \le \overline{\phi}^j\}, \label{eq:fe:constraintsets} \\
\mathcal{U}^j = \{v \in \mathcal{V}^j \,:\, \underline{\chi}^j \le v \le \overline{\chi}^j\}, \notag
\end{align}
are the \emph{downward} and \emph{upward constraint subsets} of the FE subspaces $\mathcal{V}^j$, respectively.  Observe that the upward sets are larger, $\mathcal{D}^j \subset \mathcal{U}^j$, and in fact by \eqref{eq:fe:telescoping} the set decompositions
\begin{equation}
\mathcal{U}^j = \sum_{i=0}^j \mathcal{D}^i \label{eq:fe:constraintdecomp}
\end{equation}
apply in the same sense as \eqref{eq:constraintdecomp}.  Also note that the finest-level upward constraint set is equivalent to the original constraint set $\mathcal{K}^J$; for all $z^J \in \mathcal{V}^J$,
\begin{equation}
z^J \in \mathcal{U}^J \quad \iff \quad w^J+z^J \in \mathcal{K}^J. \label{eq:fe:finestlevelequivalent}
\end{equation}
